Software Product Line Analysis Techniques

### 2018-03-07 doc/quals/quals.tex
final version of the quals document

This report covers some research work attempting to solve this problem. One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques~\cite{Apel:2013}. Another extends this work to more elaborate sampling techniques and more general categories of analysis~\cite{Liebig:2013}. At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs~\cite{Midtgaard:2015}. At the other end, some papers provide pragmatic implementations of type checkers~\cite{Kastner:2012}, source code syntax checkers~\cite{Gazzillo:2012}, data flow analyses~\cite{Bodden:2013}, and model checkers~\cite{Classen:2013}. Table \ref{tbl1} summarizes the papers covered, with limitations explained throughout this report.

@@

The evaluation presented in the paper justifies the hypotheses of the authors: the implemented variability-aware analyses under-perform single configuration analyses, but outperform the analysis of pair-wise samples. In addition, in some cases variability-aware analyses also outperform coverage-based analysis. More importantly, variability-aware analyses are complete in terms of covering all valid feature combinations (i.e. all valid products), while all the sampling techniques are incomplete in that regard.

@@

The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away). Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration. The speedup due to lifting with sharing also grows exponentially, but at a slower rate. Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only four source code methods).

@@

K\"{a}stner et a.~\cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq~\cite{Bertot:2010} proof assistant. Featherweight Java (FJ) is a functional subset of Java, leaving out constructs such as assignment (mutation) and interfaces. CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE. This type checker is further extended to fully cover Java constructs.

A product line is considered well-typed only if all the products that can be generated out of it are well-typed (typing is preserved by product generation). The type checker here takes the source code of a product line as input, and checks all variants for type errors. In addition to typing preservation, a second goal is backward compatibility. No new syntactic constructs were introduced, which allowed the use of existing tools.

One fundamental limitation of this work is that it severely restricts the allowed forms of variability. First, only disciplined annotations are supported. Although this does not restrict the expressivity of variability, it limits the applicability of this type checker to existing product lines that use undisciplined annotations. More importantly though, variability annotation is only supported within elements of syntactic sequences. For example, a sequence of methods (potentially empty) is defined. Each element (method) within that sequence can be annotated individually. However, variability is not supported for singleton syntactic constructs. This means that we cannot vary the subtype of a class or the return type of a method across products.

@@

The type checker has been evaluated on 5 Java product lines, ranging from 9 to 42 features. Some type errors were found by the checker in some product variants. The performance (in terms of time taken to type check the product line) was evaluated compared to approximate per-product performance, and on these 5 product lines processing time grew much slower than the number of variants. It is worth noting though that these product lines are relatively small, both in terms of number of lines of code and number of features.

@@

The idea behind FMLR parsing is to fork sub-parsers (initially sharing the state of their parent) on each variability point. A naive implementation would result in a number of sub-parsers exponential in the number of variability features. To prevent this exponential explosion, SuperC implements four optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces. With those optimizations, for almost all the Linux kernel compilation units the number of sub-parsers is less than 20.

@@

\section{SPLLift}

Bodden et al.~\cite{Bodden:2013} extend inter-procedural data flow analysis (based on the IFDS~\cite{Reps:1995} framework) to annotative product lines. In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems. The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions. This fits well within the IFDS framework because data flow properties are boolean expressions.

@@

Properties over FTSs are specified in fLTL, which associates presence conditions to LTL formulas. Now a variability-aware model checker needs to check an fLTL property against an FTS model. The checker returns true if the property holds in the set of products defined by the property's presence condition. If the property is violated in any of those products, a set of counterexample/presence condition pairs is returned. Each counterexample covers a set of products in which the property is violated, and each product in which the property is violated  has to be covered by a counterexample.

@@

The SNIP model checker implements FTS model checking. Its input language is fPromela, which extends Promela (used in the SPIN model checker) by feature expression guard annotation over statements. It also implements variability-aware vacuity detection and deadlock detection.

One shortcoming of this paper is that empirical evaluation is done on only one model (the mine pump controller model). This model is small compared to realistic models (only 457 FTS states), and the evaluation is done checking only 6 properties. With only one model (moreover, a small one), the evaluation results cannot be safely generalized.

@@

Late splitting and early joining are two principles that underly many of the techniques presented. These two principles maximize reuse of intermediate results calculated as a part of the analysis process. Local variability representation is another principle that manages intermediate value variability. These principles together with some analysis-specific design decisions can effectively lead to an efficient family-based analysis.
\bibliographystyle{abbrv}
\bibliography{quals} 
\end{document}

>>>>

This report covers some research work attempting to solve this problem. One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques~\cite{Apel:2013}. Another extends this work to more elaborate sampling techniques and more general categories of analysis~\cite{Liebig:2013}. At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation analyses to SPLs~\cite{Midtgaard:2015}. At the other end, some papers provide pragmatic implementations of type checkers~\cite{Kastner:2012}, source code syntax checkers~\cite{Gazzillo:2012}, data flow analyses~\cite{Bodden:2013}, and model checkers~\cite{Classen:2013}. Table \ref{tbl1} summarizes the papers covered, with limitations explained throughout this report.

@@

The evaluation presented in the paper justifies the hypotheses of the authors: the implemented variability-aware analyses under-perform single configuration analyses, but outperform the analysis of pair-wise samples. In addition, in some cases variability-aware analyses also outperform coverage-based sampling. More importantly, variability-aware analyses are complete in terms of covering all valid feature combinations (i.e. all valid products), while all the sampling techniques are incomplete in that regard.

@@

The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away). Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction is exponential with respect to the number of configurations (on the expense of accuracy), and almost identical to the average analysis performance of a single configuration. The speedup due to lifting with sharing is also exponential, but at a slower rate. Generalizing from those performance numbers is risky though because of the highly restricted sample space (a single analysis on only four source code methods).

@@

K\"{a}stner et al.~\cite{Kastner:2012} formalize and certify a Colored Featherweight Java (CFJ) type checker using the Coq~\cite{Bertot:2010} proof assistant. Featherweight Java (FJ) is a functional subset of Java, leaving out constructs such as assignment (mutation) and interfaces. CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE. This type checker is then extended to fully cover Java constructs.

An SPL is considered well-typed only if all the products that can be generated out of it are well-typed (typing is preserved by product generation). The type checker here takes the source code of a product line as input, and checks all variants for type errors. In addition to typing preservation, a secondary goal is backward compatibility with existing tools. To achieve the backward compatibility goal, no new syntactic constructs were introduced.

One fundamental limitation of this work is that it severely restricts the allowed forms of variability. First, only disciplined annotations are supported. Although this does not restrict the expressivity of variability, it limits the applicability of this type checker to existing product lines that use undisciplined annotations. More importantly though, variability annotation is only supported within elements of syntactic sequences. For example, when a sequence of methods (potentially empty) is defined, each element (method) within that sequence can be annotated individually. However, variability is not supported for singleton syntactic constructs. This means that we cannot vary the subtype of a class or the return type of a method (among many other things) across products.

@@

The type checker has been evaluated on 5 Java product lines, ranging from 9 to 42 features. Some type errors were found by the checker in some product variants. Performance (in terms of time taken to type check the product line) was evaluated compared to approximate per-product performance, and on those 5 product lines processing time grew much slower than the number of variants. It is worth noting though that those product lines are relatively small, both in terms of number of lines of code and number of features.

@@

The idea behind FMLR parsing is to fork sub-parsers (initially sharing the state of their parent) on each variability point. A naive implementation would result in a number of sub-parsers exponential in the number of variability features. To prevent this exponential explosion, SuperC implements four optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces. With those optimizations, for almost each of the Linux kernel compilation units the number of sub-parsers is less than 20.

@@

\newcommand{\spllift}{SPL\textsuperscript{Lift}}

\section{\spllift}

Bodden et al.~\cite{Bodden:2013} extend inter-procedural data flow analysis (based on the IFDS~\cite{Reps:1995} framework) to annotative product lines. In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems. The main idea behind lifting IFDS analyses is to guard data-flow properties of program statements by presence conditions. This fits well within the IFDS framework because data flow properties are boolean expressions.

@@

Properties over FTSs are specified in fLTL, which associates presence conditions to LTL formulas. Now a variability-aware model checker needs to check an fLTL property against an FTS model. The checker returns true if the property holds in the set of products defined by the property's presence condition. If the property is violated in any of those products, a set of (counterexample, presence condition) pairs is returned. Each counterexample covers a set of products in which the property is violated, and each product in which the property is violated  has to be covered by a counterexample.

@@

The SNIP model checker implements FTS model checking. Its input language is fPromela, which extends Promela (used in the SPIN model checker) with feature expression guard annotation over statements. It also implements variability-aware vacuity detection and deadlock detection.

One shortcoming of this paper is that empirical evaluation is done on only one model (the mine pump controller model). This model is small compared to realistic models (only 457 FTS states), and the evaluation is done checking only 6 properties.

@@

Late splitting and early joining are two principles that underly many of the techniques presented. These two principles maximize reuse of intermediate results calculated as a part of the analysis process. Local variability representation is another principle that manages intermediate value variability. These principles together, in addition to some analysis-specific design decisions, can effectively lead to an efficient family-based analysis of an SPL.
\bibliographystyle{abbrv}
\bibliography{quals} 
\end{document}
